Explicit Scale Simulation for analysis of RNA-sequencing with ALDEx2

### 2024-08-22 go3-author.tex
with Justin's suggestions
Send back to Justin for comment

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Normalizations are not what you think; Explicit Scale Simulation in ALDEx2},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Normalizations are not what you think; Explicit Scale Simulation
in ALDEx2}
\author{Gregory B. Gloor$^1$ \and Michelle Pistner Nixon$^2$ \and Justin D. Silverman$^{2,3,4}$}
\date{\today}

@@

\begin{abstract}
In high-throughput sequencing (HTS) studies, sample-to-sample variation
in sequencing depth is driven by technical factors, and not by variation
in the scale (e.g., total size, microbial load, or total gene
expression) of the underlying biological systems. Typically a
statistical normalization is used to remove unwanted technical variation
in the data or the parameters of the model to enable analyses that are
sensitive to scale; e.g., differential abundance and differential
expression analyses. Recently we showed that all normalizations make
implicit assumptions about the unmeasured system scale and that errors
in these assumptions can lead to dramatic increases in false positive
and false negative rates. Here we describe updates to the ALDEx2 R
package that mitigate these problems by directly modeling uncertainty in
the unmeasured system scale through the use of a \textit{scale model}.
Scale models generalize the idea of normalizations and can be thought of
as explicitly modeling the error in normalization by examining a
distribution over all possible normalizations. Beyond enhancing the
robustness of HTS analyses, the use of scale models within ALDEx2
enhances the transparency and reproducibility of analyses by making
implicit normalizing assumptions an explicit part of the model building
process.
\end{abstract}

@@

High-throughput sequencing (HTS) is a ubiquitous tool used to explore
many biological phenomenon such as gene expression (single-cell
sequencing, RNA-sequencing, meta-transcriptomics), microbial community
composition (16S rRNA gene sequencing, shotgun metagenomics) and
differential enzyme activity (selex, CRISPR killing). HTS proceeds by
taking a sample from the environment, making a library, multiplexing
(merging) multiple libraries together, and then applying a sample of the
multiplexed library to the flow cell. Each of these steps is a
compositional sampling step as only a fixed-size subsample of nucleic
acid is carried over to subsequent steps. Thus, with each sampling step
the connection between the size of the sampled DNA pool and the scale
(e.g., size, microbial load, or total gene expression) of the measured
biological system is degraded or lost. Increasingly, researchers are
turning to modified experimental protocols (e.g., DNA spike-ins or cell
counting) in an attempt to recover the lost biological variation in
scale; see for example (1, 2). However, these protocols often do not
recover meaningful biological variation but instead recover scale
variation at the intermediate step in the sample preparation protocol
where the DNA spike-in was added. In short, the disconnect between
sample-to-sample variation in sequencing depth and biological variation
in scale remains an outstanding challenge.

The analysis of HTS data suffers from several known problems that can be
traced, in whole or in part, to misspecification of scale. The first
issue is poor control of the false discovery rate (FDR)
{[}(3);(4);Nearing:\href{mailto:2022aa;@Li}{\nolinkurl{2022aa;@Li}}:2022aa{]},
exhibited as dataset-dependent FDR control. The FDR problem is connected
to the double-filtering approach that is often used, but which is known
not to have appropriate FDR control (5, 6). The second issue is poor
performance when analyzing asymmetric data; data that arises because of
a bias in the direction of change. This type of data frequently arises
in in-vitro selection experiments (SELEX), transcriptome analysis, and
microbiome analysis (7). The third issue is that these problems become
more pronounced as more samples are collected; that is, more information
results in a worsening of the accuracy of the analysis (9). The final
problem is that these data are relative data, i.e.~compositional (10,
11), and substantial effort has been put into removing this constraint.

The first three problems were recently shown by Nixon and Silverman (8)
to be a result of a mismatch between the underlying size or scale of the
system and the assumptions of the normalizations used for the analysis
of HTS. Biological variation in scale often represents an important
unmeasured confounder in HTS analyses (12). For example, cells
transformed by the cMyc oncogene have about 3 times the amount of mRNA
and about twice the rRNA content than non-transformed cells (13), and
this dramatically skews transcriptome analysis (14). In addition,
wild-type and mutant strains of cell lines, yeast or bacteria have
different growth rates and RNA contents under different conditions,
which affect our ability to identify truly differentially abundant genes
(15--17). As another example, the total bacterial load of the vaginal
microbiome differs by 1-2 orders of magnitude in absolute abundance
between the healthy and bacterial vaginosis states (18), and the
composition between these states is dramatically different (19, 20).
Thus, a full description of any of these systems includes both relative
change (composition) and absolute abundance (scale). Current methods
access only the compositional information yet make implicit assumptions
about the scale.

Recently, Nixon et al. (8) showed that the challenge of non-biological
variation in sequencing depth be viewed as a problem of
partially-identified models. They showed that \emph{all} normalizations
make some assumption about scale. These implicit assumptions are often
difficult to interpret, and different normalizations provide different
outputs when applied to the same dataset (3, 21--24). Intuitively,
normalizations in widespread use assume that either all samples have the
same scale, e.g.~proportions, rarefaction (25), RPKM (26, 27), etc; or
that a subset of features in one sample can be chosen as a reference to
which the others are scaled e.g.~the TMM (28), or LVHA (7) or the
additive log-ratio (29); or that different sub-parts of each sample
maintain a constant scale across samples e.g.~the RLE (30); or that the
geometric mean of the parts is appropriate e.g the CLR and its
derivatives.

The original ALDEx2 (31) model unwittingly made a strict assumption
about scale through the CLR normalization (8). While this assumption
could be useful in many cases it could never be exactly true, and others
have shown that it is not always the best option (32). Nixon et al. (8)
showed that better scale assumptions resulted in more reproducible data
analysis including better control of both false positive and false
negative results. In essence ALDEx2 has been modified to explicitly
model the scale over a range of reasonable normalization parameters.
Here, we briefly introduce these modifications and then show how scale
uncertainty can greatly improve modeling in transcriptome and
meta-transcriptome datasets to provide more robust and reproducible
results.

@@

To be concrete, we let \(\mathbf{Y}\) denote the \emph{measured}
\(D \times N\) matrix of sequence counts with elements
\(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules
mapping to feature \(d\) (e.g., a taxon or gene) in sample \(n\).
Likewise, we can denote \(\mathbf{W}\) as the \emph{true} amount of
class \(d\) in the biological system from which sample \(n\) was
obtained. We can think of \(\mathbf{W}\) as consisting of two parts, the
scale \(\mathbf{W}^{\perp}\) (e.g., totals) and the composition
\(\mathbf{W}^{\parallel}\) (i.e., proportions). That is,
\(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements
\(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while
\(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements
\(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\).
Note that with these definitions \(\mathbf{W}\) can be written as the
element-wise combination of scale and composition:
\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).

All the normalizations in current use are data transformations that can
be stated as ratios of the form
\(\hat{{\mathbf{W}}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\), where the
denominator is determined by some function of the observation. We use
the \(\hat{{\mathbf{W}}}\) notation to indicate that the normalization
is attempting to provide an estimate of the true value. The technical
variation in sequencing depth
(\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that
observed data \(\mathbf{Y}\) provides us with information about the
system composition \(\mathbf{W}^{\parallel}\) but little to no
information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et
al.~2011).

@@

The ALDEx2 R package (31, 33) is a general purpose toolbox for Bayesian
modeling of HTS data. For brevity, we discuss ALDEx2 in its simplest
form as a tool for estimating the magnitude and statistical significance
of Log-Fold-Changes (LFC) (e.g., differential abundance or differential
expression analysis), but note that it can be used to fit more complex
linear models. At a high-level, ALDEx2 involves three steps: 1) a
Bayesian model is used to estimate \(\mathbf{\hat{W}}^{\parallel}\)
given observations \(\mathbf{Y}\); 2) a normalization is used to
estimate \(\mathbf{\hat{W}}\) given the estimate
\(\mathbf{\hat{W}}^{\parallel}\); 3) the \(\mathbf{\hat{W}}\) estimates
are used to estimate LFC as part of differential abundance (or
expression) analyses. For more details on ALDEx2 see (8, 31).

By default, ALDEx2 uses the CLR normalization which can be written as
\(\log \mathbf{\hat{W}}_{dn}=\log (\mathbf{\hat{W}}^{\parallel}_{dn}/G_{n})\)
where \(G_{n}\) denotes the geometric mean of the vector
\((\hat{W}^{\parallel}_{1n}, \dots, \hat{W}^{\parallel}_{Dn})\). The CLR
normalization equates to an implicit assumption that
\(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\)(8) and even slight errors in
this assumption introduces bias into the LFC estimate that is not
accounted for when estimating uncertainty (e.g., confidence intervals or
credible sets). In fact, Nixon et al. (8) showed that the only way in
which the ALDEx2 model, or any normalization-based model, could ever be
calibrated (e.g., control Type-I Error rates) was if this assumption was
exactly true. In support of this assertion it is known that false
discovery rates vary widely by analysis, dataset, and normalization
method (34, 35). To fix this, Nixon et al. (8) showed that models should
incorporate potential error in the assumptions implied by
normalizations.

Nixon et al.~(2023) generalized the concept of normalizations by
introducing the concept of a \textit{scale model} to account for
potential error. Scale models can be incorporated into ALDEx2, turning
the ALDEx2 model into a specialized type of statistical model which they
called a \textit{Scale Simulation Random   Variable} (SSRV). They did
this by including a model for \(\mathbf{\hat{W}}^{\perp}_{n}\). Since
the CLR normalization makes the assumption
\(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), the CLR normalization can be
generalized by considering probability models for the scale
\(\mathbf{\hat{W}}^{\perp}_{n}\) that have mean \(1/G_{n}\). For
example, the following scale model generalizes the CLR:

\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(0, \gamma^{2})\]

where \(\gamma\) is a tunable parameter drawn from a Gaussian
distribution that controls the degree of uncertainty in the CLR
assumption and \(x_{n}\) denotes a binary condition indicator (e.g.,
\(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control). We have made
those modifications a permanent fixture of ALDEx2 which now represents
the first software package designed for SSRV-based inference.

@@

It is standard practice in many fields of HTS, but especially
transcriptomics, to use the dual cutoff approach graphically exemplified
by volcano plots (36, 37) because not all statistically significant
differences are biologically relevant. Paraphrasing this we can say that
in in some types of datasets more samples leads to a majority of
features being statistically significant. A detailed reasoning for this
using modelled data is explained in two reports by Nixon et al. (8, 9),
which shows that scaled models completely address this analytic issue.

We use the data of Gierliński et al. (38) who conducted a highly
replicated yeast transcriptome experiment comparing a wild-type strain
with a snf2 gene knockout, \(\Delta\)snf2. This dataset has been used to
argue that a dual cutoff approach is appropriate to limit the number of
significant parts and for purposes of reproducibility (37). However, in
this benchmarking study, the number of significantly different
transcripts varied between 65\% to \textgreater80\% of all transcripts
depending on the tool; in essence the desired behavior was that almost
all transcripts were significantly different. The guidance on
reproducibility runs counter to standard statistical practice where
power is intrinsically linked to sample size and very large sample sizes
are indeed desired (39). Furthermore, the dual-cutoff approach is known
not to provide appropriate FDR control (5, 6). Schurch et al. (37)
further suggested that different tools might be better in some
conditions or datasets than others because each tool has different
intrinsic statistical power and Type 1 and Type 2 errors. Through the
lens of scale uncertainty, the behaviour of the tools in this study show
unacknowledged bias; false confidence in the precision of the estimate
as sample size increases because o mis-specification in the scale of the
data. This certainty is driven by the assumptions of the tool not the
actual experiment being investigated (9).

Using either DESeq2 or ALDEx2, a majority of transcripts are
statistically significantly different between groups with a
Benjamini-Hochberg (40) false discovery rate (FDR) of 0.05; i.e.~4636
(79\%, DESeq2) or 4172 (71\%, ALDEx2) of the 5891 transcripts. Such
large numbers of significant transcripts seems biologically unrealistic
and furthermore breaks the necessary assumption made by the
normalization methods that the majority of the features must be
invariant. That 118 transcripts are identified by ALDEx2 and not DESeq2,
while DESeq2 identifies 582 transcripts that ALDEx2 does not, suggests
that the choice of normalization plays a role in which results are
returned as significant and that some if not the majority are driven by
technical differences in the analysis (22, 23, 37, 41, 42).

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Explicit Scale Simulation for analysis of RNA-sequencing with ALDEx2},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Explicit Scale Simulation for analysis of RNA-sequencing with
ALDEx2}
\author{Gregory B. Gloor$^1$ \and Michelle Pistner Nixon$^2$ \and Justin D. Silverman$^{2,3,4}$}
\date{\today}

@@

\begin{abstract}
In high-throughput sequencing (HTS) studies, sample-to-sample variation
in sequencing depth is driven by technical factors, and not by variation
in the scale (e.g., total size, microbial load, or total mRNA
expression) of the underlying biological systems. Typically a
statistical normalization is used to remove unwanted technical variation
in the data or the parameters of the model to enable analyses that are
sensitive to scale; e.g., differential abundance and differential
expression analyses. We recently showed that all normalizations make
implicit assumptions about the unmeasured system scale and that errors
in these assumptions can dramatically increase false positive and false
negative rates. We demonstrated that these errors can be mitigated by
accounting for uncertainty about scale using a \emph{scale model}, which
we integrated into the ALDEx2 R package. This article provides new
insights into those methods, focusing on the application to
transcriptomic analysis. Here we provide transcriptomic case studies
demonstrating how scale models, rather than traditional normalizations,
can reduce false positive and false negative rates in practice while
enhancing the transparency and reproducibility of analyses. We show that
these scale models replace the need for dual cutoff approaches often
used to address the disconnect between practical and statistical
significance. We demonstrate the utility of that scale models built
based on known housekeeping genes in complex metatranscriptomic
datasets. Thus this work provides example and practical guidance on how
to incorporate scale into transcriptomic analysis.
\end{abstract}

@@

\doublespacing

High-throughput sequencing (HTS) is a ubiquitous tool used to explore
many biological phenomenon such as gene expression (single-cell
sequencing, RNA-sequencing, meta-transcriptomics), microbial community
composition (16S rRNA gene sequencing, shotgun metagenomics) and
differential enzyme activity (selex, CRISPR killing). HTS proceeds by
taking a sample from the environment, making a library, multiplexing
(merging) multiple libraries together, and then applying a sample of the
multiplexed library to a flow cell. Each of these steps is a
compositional sampling step as only a fixed-size subsample of nucleic
acid is carried over to subsequent steps. Thus, with each sampling step
the connection between the size of the sampled DNA pool and the scale
(e.g., size, microbial load, or total gene expression) of the measured
biological system is degraded or lost. In the end, the information
contained in the data relates only to relative abundances and has an
arbitrary scale imposed by the sequencing process(1--3). Increasingly,
researchers are turning to modified experimental protocols in an attempt
to recover the lost biological variation in scale; see for example (4,
5). However, these protocols only recover scale variation at the
intermediate step in the sample preparation protocol where the
intervention was made. In short, the disconnect between sample-to-sample
variation in sequencing depth and biological variation in scale remains
an open challenge.

The analysis of HTS data suffers from several known problems that can be
traced, in whole or in part, to misspecification of scale. The first
issue is poor control of the false discovery rate (FDR) (6--9),
exhibited as dataset-dependent FDR control and by the disconnect between
statistical and biological significance(10). In current practice, these
issues are addressed by a dual-filtering method, whereby both a low
p-value (or equivalently a low q-value following FDR correction (11))
and a large difference between groups is used to identify interesting
transcripts or genes for follow-up analysis(10). This double-filtering
approach is graphically exemplified by the volcano plot (12), but is
known to not appropriately control the FDR\\
(13, 14). The second issue is poor performance when analyzing data where
the mean change between groups is non-zero. Such asymmetric data can
arise when a gene set is expressed in one group but not the other, or
when one group contains different gene content from the other. This type
of data frequently arises in in-vitro selection experiments (SELEX),
transcriptome analysis, and microbiome analysis (15). The third issue is
that these problems become more pronounced as more samples are
collected; that is, more information results in a worsening of the
accuracy of the analysis (16).

The three problems were recently shown by Nixon and Silverman (3) to be
a result of a mismatch between the underlying size or scale of the
system and the assumptions of the normalizations used for the analysis
of HTS. Biological variation in scale often represents an important
unmeasured confounder in HTS analyses (17). For example, cells
transformed by the cMyc oncogene have about 3 times the amount of mRNA
and about twice the rRNA content than non-transformed cells (18), and
this dramatically skews transcriptome analysis (19). In addition,
wild-type and mutant strains of cell lines, yeast or bacteria have
different growth rates and RNA contents under different conditions,
which affect our ability to identify truly differentially abundant genes
(20--22). As another example, the total bacterial load of the vaginal
microbiome differs by 1-2 orders of magnitude in absolute abundance
between the healthy and bacterial vaginosis states (23), and the
composition between these states is dramatically different (24, 25).
Thus, a full description of any of these systems includes both relative
change (composition) and absolute abundance (scale). Current methods
access only the compositional information yet make implicit assumptions
about the scale (16).

Recently, Nixon et al. (3) showed that the challenge of non-biological
variation in sequencing depth be viewed as a problem of
partially-identified models. They showed that \emph{all} normalizations
make some assumption about scale but these implicit assumptions are
often inappropriate and difficult to interpret. Because of this
different normalizations provide different outputs when applied to the
same dataset (6, 8, 26--28). Intuitively, normalizations in widespread
use assume that either all samples have the same scale,
e.g.~proportions, rarefaction (29), RPKM (30, 31), etc; or that a subset
of features in one sample can be chosen as a reference to which the
others are scaled e.g.~the TMM (32), or LVHA (15) or the additive
log-ratio (33); or that different sub-parts of each sample maintain a
constant scale across samples e.g.~the RLE (34); or that the geometric
mean of the parts is appropriate e.g the CLR (35) and its derivatives.

The original naive ALDEx2 (36) model unwittingly made a strict
assumption about scale through the CLR normalization (3). While this
assumption could be useful in many cases it could never be exactly true,
and others have shown that it is not always the best option (37). Nixon
et al. (3) showed that better scale assumptions resulted in more
reproducible data analysis including better control of both false
positive and false negative results. We recently modified ALDEx2 to
explicitly model the scale over a range of reasonable normalization
parameters, and showed significant improvements in performance in
microbiome and in-vitro selection experiments (16). Here, we briefly
introduce review modifications and then show how scale uncertainty can
greatly improve modeling in transcriptome and meta-transcriptome
datasets to provide more robust and reproducible results.

@@

Formal and expanded descriptions of the concepts that follow are given
in (3, 16). To be concrete, we let \(\mathbf{Y}\) denote the
\emph{measured} \(D \times N\) matrix of sequence counts with elements
\(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules
mapping to feature \(d\) (e.g., a taxon, transcript or gene) in sample
\(n\). Likewise, we can denote \(\mathbf{W}\) as the \emph{true} amount
of class \(d\) in the biological system from which sample \(n\) was
obtained. We can think of \(\mathbf{W}\) as consisting of two parts, the
scale \(\mathbf{W}^{\perp}\) (e.g., totals) and the composition
\(\mathbf{W}^{\parallel}\) (i.e., proportions). That is,
\(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements
\(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while
\(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements
\(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\).
Note that with these definitions \(\mathbf{W}\) can be written as the
element-wise combination of scale and composition:
\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\),
or as the logarithm
\(\log \mathbf{W}_{dn}= \log \mathbf{W}^{\parallel}_{dn} + \log \mathbf{W}^{\perp}_{n}\).

Many of the normalizations in widespread use in tools such as DESeq2
(38), edgeR (32), metagenomeSeq (39) ALDEx2 (40) can be stated as ratios
of the form
\(\hat{{\mathbf{W}}}_{dn} \approx \mathbf{Y}_{dn}/f(\mathbf{Y})\), where
the denominator is determined by some function of the observation. We
use the \(\hat{{\mathbf{W}}}\) (\(\ \hat{}\ \)) notation to indicate
that the output is an estimate of the true value. The technical
variation in sequencing depth
(\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that
observed data \(\mathbf{Y}\) provides us with information about the
system composition \(\mathbf{W}^{\parallel}\) but little to no
information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et
al.~2011).

@@

The ALDEx2 R package (36, 40) is a general purpose toolbox to model the
uncertainty of HTS data and to use that model to estimate LFC
significance. At a high-level, ALDEx2 has three connected components to
estimate the uncertainty inherent in HTS datasets. First, the tool
accounts for the uncertainty of the sequencing counts using Dirichlet
multinomial sampling to build a probabilistic model of the data; i.e.,
\(\mathbf{\hat{W}}^{\parallel} \approx \mathrm{Dir}(\mathbf{Y})\).
Secondly, ALDEx2 uses the centred log-ratio transformation to scale the
data (36). However, this step was modified recently to account for scale
uncertainty and misspecification (16) via a scale model, explained with
more details in (3, 16) and summarized in the next paragraph. Finally, a
standard null-hypothesis test and a non-parametric estimate of mean
standardized difference are used to report on the finite sample
variation. These sources of uncertainty and variation are combined via
reporting the expected values from a Monte-Carlo simulation framework.
For simplicity, we use the term `difference' to refer to the absolute
difference between groups, and `dispersion' to refer to the
within-condition difference or pooled variance as defined in (36). These
are calculated on a \(\log_2\) scale. For more details on ALDEx2 see (3,
16, 36, 40).

Scale models can be incorporated into ALDEx2, turning the ALDEx2 model
into a specialized type of statistical model which Nixon et al. (3) term
a \textit{Scale Simulation Random   Variable} (SSRV). To do this, Nixon
et al. (3) generalized the concept of normalizations by introducing the
concept of a \textit{scale model} to account for potential error in the
centred log-ratio normalization step. They did this by including a model
for \(\mathbf{\hat{W}}^{\perp}_{n}\). The CLR method used by ALDEx2
makes the assumption \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), where
\(G_n\) is the geometric mean of sample n, which while being a random
variable, is essentially constant across each Monte-Carlo replicate.
With this modificaiton, ALDEx2 can be generalized by considering
probability models for the scale \(\mathbf{\hat{W}}^{\perp}_{n}\) that
have mean \(1/G_{n}\). For example, the following scale model
generalizes the CLR:

\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(\mu, \gamma^{2})\].

This formulation is quite flexible (3, 16). In the simple or `default'
configuration, \(\mu = 0\) and \(\gamma\) is a tunable parameter drawn
from a log-Normal distribution that adds scale uncertainty. The default
approach controls only the degree of uncertainty in the CLR assumption
for the \(x_{n}\) binary condition indicator (e.g., \(x_{n}=1\) denotes
case and \(x_{n}=0\) denotes control). In the advanced or `informed'
configuration, \(\mu\) takes different values for each group, and this
controls the location of the LFC assumption. An example of both the
default and informed approaches is given for a microbiome dataset in
(16), and here we show that these approaches also work well in
transcriptome and metatranscriptome datasets. These modifications are
instantiated in ALDEx2 which is the first software package designed for
SSRV-based inference.

@@

Gierliński et al. (41) generated a highly replicated yeast transcriptome
dataset to compare gene expression between a wild-type strain and a snf2
gene knockout, \(\Delta\)snf2. This dataset was used to test several
RNA-seq tools for their power to detect the set of differentially
abundant transcripts identified in the full dataset when the data was
subset (10). In this study each tool had its own `gold standard' set of
transcripts with different tools identifying between between 65\% to
\textgreater80\% of all transcripts as being significantly different.
Since the majority of transcripts were significantly different, the
authors suggested that it was more appropriate to apply a dual cutoff
composed of both a Benjamini-Hochberg (42) corrected p-value (q-value)
plus a difference cutoff to limit the number of identified transcripts
to a much smaller fraction of the total. Nixon et al, (16) showed that
adding even a small amount of scale uncertainty with ALDEx2 dramatically
reduced the number of significant transcripts identified, removing the
need for the dual cutoff approach in this dataset and others. Below we
include an intuitive explanation of how incorporating scale uncertainty
achieved this outcome using a setting of \(\gamma=0.5\), which is in
line with the recommendations of (16) and that gives results comparable
to those proposed in (10).

We start with the assumption that not all statistically significant
differences are biologically relevant (43), and that such a large number
of significantly different transcripts breaks the necessary assumption
for DA/DE expression that most parts be invariant (27). As noted,
transcriptomics commonly uses a dual cutoff approach that is graphically
exemplified by volcano plots (10, 12). Using either DESeq2 or ALDEx2, a
majority of transcripts are statistically significantly different
between groups with a q-value cutoff of \(\le 0.05\); i.e.~4636 (79\%,
DESeq2) or 4172 (71\%, ALDEx2) of the 5891 transcripts. These values are
in line with those observed by (10). Such large numbers of statistically
significant transcripts seems biologically unrealistic. That 118
transcripts are identified by ALDEx2 and not DESeq2, while DESeq2
identifies 582 transcripts that ALDEx2 does not, suggests that the
choice of normalization plays a role in which results are returned as
significant and that some, if not the majority, are driven by technical
differences in the analysis (9, 27).
